\Large Carryover effects in free recall reveal how past experiences influence memories of future experiences

### 2023-01-17 paper/main.tex
spell check and some compilation error fixes

Participants were assigned to experimental conditions based loosely on their
date of participation. (This aspect of our procedure helped us to more easily
synchronize the experiment databases across multiple testing computers.) Of the
490 participants who opted to fill out the demographics survey, reported ages
ranged from 17 to 31 years (mean: 19.1 years; standard deviation: 1.356 years).
A total of 318 participants reported their gender as female, 170 as male, and
two participants declined to report their gender. A total of 442 participants
reported their ethnicity as ``not Hispanic or Latino,'' 39 as ``Hispanic or
Latino,'' and nine declined to report their ethnicity. Participants reported
their races as White (345 participants), Asian (120 participants), Black or
African American (31 participants), American Indian or Alaska Native (11
particiapnts), Native Hawaiian or Other Pacific Islander (four participants),
Mixed race (three participants), Middle Eastern (one participant), and Arab
(one participant). A total of five participants declined to report their race.
We note that several participants reported more than one of racial category.
Participants reported their highest degrees achieved as ``Some college'' (359
participants), ``High school graduate'' (117 participants), ``College
graduate'' (seven participants), ``Some high school'' (five participants),
``Doctorate'' (one participant), and ``Master's degree'' (one participant). A
total of 482 participants reported no reading impairments, and eight reported
having mild reading impairments. A total of 489 participants reported having
normal color vision and one participant reported that they were red-green color
blind. A total of 482 participants reported taking no prescription medications
and having no recent injuries; four participants reported having ADHD, one reported
having dyslexia, one reported having allergies, one reported a recently torn
ACL/MCL, and one reported a concussion from several months prior. The
participants reported consuming 0 -- 3 cups of coffee prior to the testing
session (mean: 0.32 cups; standard deviation: 0.58 cups). Participants reported
their current level of alertness, and we converted their responses to numerical
scores as follows: ``very sluggish'' (-2), ``a little sluggish'' (-1),
``neutral'' (0), ``a little alert'' (1), and ``very alert'' (2). Across all
participants, the full range of alertness levels were reported (range: -2 -- 2;
mean: 0.35; standard deviation: 0.89).

@@

Our experiment is a variant of the classic free recall paradigm that we term
\textit{feature-rich free recall}. In feature-rich free recall, participants
study 16 lists, each comprised of 16 words that vary along a number of stimulus
dimensions (Fig.~\ref{fig:exp}). The stimulus dimensions include two semantic
features related to the \textit{meanings} of the words (semantic category,
referrent object size), two lexicographic features related to the
\textit{letters} that make up the words (word length in number of letters,
identity of the word's first letter), and two visual features that are
independent of the words themselves (text color, presentation location). Each
list contains four words from each of four different semantic categories and
two object sizes; all other stimulus features are randomized. After studying
each list, the participant attempts to recall as many words as they can from
that list, in any order they choose. Because each individual word is associated
with several well-defined (and quantifiable) features, and because each list
incorporates a diverse mix of feature values along each dimension, this allows
us to estimate which features participants are considering or leveraging in
organizing their memories.

@@

We assigned the categorized words into a total of 16 lists with several
constraints. First, we required that each list contained words from exactly 4
unique categories, each with exactly 4 examplars from each category. Second, we
required that (across all words on the list) at least one instance of both
object sizes were represented. On average, each category was represented in
4.27 lists (standard deviation: 1.16 lists). Aside from these two constraints,
we assigned each word to a unique list. After random assignment, each list
contained words with an average of 11.13 unique starting letters (standard
deviation: 1.15 letters) and an average word length of 6.17 letters (standard
deviation: 0.34 letters).

@@

To select a random font color for each word, we drew three integers uniformly
and at random from the interval $\[0, 255\]$, corresponding to the red (r),
green (g), and blue (b) color channels for that word. To assign random
presentation locations to each word, we selected two floating point numbers
uniformly at random (one for the word's horizontal $x$ coordinate and the other
for its vertical $y$ coordinate). The bounds of these coordinates were selected
to cover the entire visible area of the display without cutting off any part of
the words. The words were shown on 27~in (diagonal) Retina 5K iMac displays
(resolution: 5120 $\times$ 2880 pixels).

@@

We used four ``control'' conditions to evaluate and explore participants'
baseline behaviors. We also used performance on these control conditions to
help interpret performance in other ``manipulation'' conditions. Two control
conditions served as ``anchorpoints.'' In the first anchorpoint condition,
which we call the \textit{feature rich} condition, we randomly shuffled the
presentation order (independently for each participant) of the words on each
list. In the second anchorpoint condition, which we call the \textit{reduced}
condition, we randomized word presentations as in the feature rich condition.
However, rather than assigning each word a random color and location, we
instead displayed all of the words in black and at the center of the screen.

@@

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique lists for each participant.
First, we choose a word uniformly at random from the set of candidates. Next,
we compute the distances between the chosen word's feature(s) and the
corresponding feature(s) of all yet-to-be-presented words. Third, we convert
these distances (between the previously presented word's feature values, $a$,
and the candidate word's feature values, $b$) to similarity scores:
\begin{equation} \mathrm{similarity}(a, b) = \exp\{-\tau \cdot
\mathrm{distance}(a, b)\}, \label{eqn:similarity} \end{equation} where $\tau =
1$ in our implementation. We note that increasing the value of $\tau$ would
amplify the influence of similarity on order, and decreasing the value of
$\tau$ would diminish the influence of similarity on order. Also note that this
approach requires $\tau > 0$. Finally, we computed a set of normalized
similarity values by dividing the similarities by their sum: \begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, \end{equation} where in the
demoniator, $i$ takes on each of the $n$ feature values of the to-be-presented
words. The resulting set of normalized similarity scores sums to one.

@@

\caption{\textbf{Generating stochastic feature-sorted lists.} For a given
feature dimension (e.g., color), we compute the similarity
(Eqn.~\ref{eqn:similarity}) between the feature value(s) of the previous item,
$x$, and all yet-to-be-presented items ($a$ -- $g$). Next, we normalize these
similarity scores so that they sum to one. We lay, in sequence, a set of
``sticks,'' one for each candidate item, whose lengths are equal to these
normalized similarity scores. Note that the combined lengths of these sticks is
one. To select the next to-be-presented item, we draw a random number, $r$,
from the uniform distibution bounded between 0 and 1 (inclusive). The identity
of the next item is given by the stick adjacent to an indicator that moves
distance $r$ (starting from 0) along the sequence of sticks. In this case, the
next to-be-presented item is $e$. Note that each item's chances of selection is
proportional to its similarity to the previous item, along the given feature
dimension.} \label{fig:stick}

@@

Suppose that two lists contain unequal numbers of items of each size. For
example, suppose that list $A$ contains all ``large'' items, whereas list $B$
contains an equal mix of ``large'' and ``small'' items. For a participant
recalling list $A$, any correctly recalled item will necessarily match the size
of the previous correctly recalled item. In other words, successively recalling
several list $A$ items of the same size is essentially meaningless, since
\textit{any} correctly recalled list $A$ word will be large. In contrast,
successively recalling several list $B$ items \textit{could} be meaningful,
since (early in the recall sequence) the yet-to-be-recalled items come from a
mix of sizes. However, once all of the small items on list $B$ have been
recalled, the best possible next matching recall will be a large item. And all
subsequent correct recalls must also be large items-- so for those later
recalls it becomes difficult to determine whether the participant is
successively recalling large items because they are organizing their
memories according to size, or (alternatively), whether they are simply
recalling the yet-to-be-recalled items in a random order. In general, the
precise order and blend of feature values expressed in a given list, the orders
and numbers of correct recalls a participant makes, the number of itervening
presentation positions between successive recalls, and so on, can all affect
the range of clustering scores that are possible to observe for a given list.
An uncorrected clustering score therefore conflates participants' actual
memory organization with other ``nuisance'' factors.

@@

We define each participant's \textit{memory fingerprint} as the set of their
permutation-corrected clustering scores across all dimensions we tracked in our
study, including their six feature-based clustering scores (category, size,
length, first letter, color, and location) and their temporal clustering score.
Conceptually, a participant's memory fingerprint describes their tendancy to
order in their recall sequences (and, presumably, organize in memory) the
studied words along each dimension. To obtain stable estimates of these
fingerprints for each participant, we averaged clustering scores across lists.
We also tracked and characterized how participants' fingerprints changed across
lists (e.g.,
Figs.~\ref{fig:fingerprint-trajectories},~\fingerprintTrajectoryRandom).

@@

The presentation orders of some lists in the adaptive condition of our
experiment (see \textit{Adaptive condition}) were sorted according to
participants' \textit{current} memory fingerprint, estimated using all of the
lists they had studied up to that point in the experiment. Because our
experiment incorporated a speech-to-text component, all of the behavioral data
for each participant could be analyzed just a few seconds after the conclusion
of the recall intervals for each list. We used the \texttt{Quail} Python
package~\citep{HeusEtal17} to apply speech-to-text alorithms to the
just-collected data, aggregate the data for the given participant, and estimate
the participant's memory fingerprint using all of their available data up to
that point in the experiment. Two aspects of our implementation are worth
noting. First, because memory fingerprints are computed independently for each
list and then averaged across lists, the already-computed memory fingerprints
for earlier lists could be cached and loaded as needed in future computations.
This meant that our computations pertaining to updating our estimate of a
participant's memory fingerprint only needed to consider data from the most
recent list. Second, each element of the null distributions of uncorrected
fingerprint scores (see \textit{Permutation-corrected feature clustering
scores}) could be estimated independently from the others. This enabled us to
make use of the testing computers' multi-core CPU architectures by elements of
the null distributions in batches of eight (i.e., the number of CPU cores on
each testing computer). Taken together, we were able to compress the relevant
computations into just a few seconds of computing time. The combined processing
time for the speech-to-text algorithm, fingerprint computations, and
permutation-based ordering procedure (described next) easily fit within the
inter-list intervals, where participants paused for a self-paced break before
moving on to study and recall the next list.

@@

Probability of first recall curves~\citep{AtkiShif68, PostPhil65, WelcBurn24}
reflect the probability that an item will be recalled first, as a function of
its serial position during encoding. To carry out this analysis, we initialized
(for each participant) a number-of-lists (16) by number-of-words-per-list (16)
matrix of zeros. Then, for each list, we found the index of the word that was
recalled first, and we filled in that position in the matrix with a 1. Finally,
we averaged over the rows of the matrix to obtain a 1 by 16 array of
probabilities, for each participant. We used an analogous procedure to compute
probabilility of $n$\textsuperscript{th} reacall curves for each participant.
Specifically, we filled in the corresponding matrices according to the
$n$\textsuperscript{th} recall on each list that each participant made. When a
given participant had made fewer than $n$ recalls for a given list, we simply
excluded that list from our analysis when computing that paritcipant's
curve(s).

@@

We used the distances between feature values for successively presented words
(see \textit{Defining feature-based distances}) to estimate ``event
boundaries'' where the feature values changed more than
usual~\citep{EzzyDava11, MannEtal16,RadvCope06, SwalEtal09, SwalEtal11,
DuBrDava16}. For each list, for each feature dimension, we computed the
distribution of distances between the feature values for successively presented
words. We defined event boundaries (e.g., Fig.~\ref{fig:recall-dynamics}B) as
occuring between any successive pair of words whose distances along the given
feature dimension were greater than one standard deviation above the mean for
that list. Note that, because event boundaries are defined for each feature
dimension, each individual list may contain several sets of event boundaries,
each at different moments in the presentation sequence (depending on the
feature dimension of interest).

@@

While holding the set of words (and the assignments of words to lists)
constant, we manipulated two aspects of participants' experiences of studying
each list. We sought to understand the effects of these manipulations on
participants' memories for the studied words. First, we added two additional
sources of visual variation to the individual word presentations: font color
and onscreen location. Importantly, these visual features were independent of
the meaning or semantic content of the words (e.g., word category, size of the
refferent, etc.) and of the lexicographic properties of the words (e.g., word length,
first letter, etc.). We wondered whether this additional word-independent information
might facilitate recall (e.g., by providing new potential ways of organizing or
retrieving memories of the studied words) or impair recall (e.g., by
distracting participants with irrelevant information). Second, we manipulated
the orders in which words were studied (and how those orderings changed over
time). We wondered whether presenting the same list of words with different
appearances (e.g., by manipulating font size and onscreen location) or in
different orders (e.g., sorted along one feature dimension versus another)
might serve to influence how participants organized their memories of the
words. We also wondered whether some order manipulations might be temporally
``sticky'' by influencing how \textit{future} lists were remembered.

@@

manipulation conditions was indistinguishable from performnace on the early

@@

the semanticly ordered conditions exhibited stronger temporal clustering on

@@

conditions. Since results were similar across both semantic conditoins

@@

When we closely examined the sequences of words participants recalled in early
order manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Figs.~\dynamicsRandom,~\recallInit). One striking difference is that
participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}B; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to display an
icnreased tendancy to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncracies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing irrelevant visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

@@

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. We
defined participants' \textit{memory fingerprints} as the set of temporal and
feature clustering scores. A participant's memory fingerprint describes how
they tend to retrieve memories of the studied items, perhaps searching through
several feature spaces (or along several representational dimensions). To gain
insights into the dynamics of how participants' clustering scores tended to
change over time, we computed the average (across participants) fingerprint
from each list, from each order manipulation condition
(Fig.~\ref{fig:fingerprint-trajectories}). We projected these fingerprints into
a two-dimensional space to help visualize the dynamics (top panels; see
\textit{Computing low-dimensional embeddings of memory fingerprints}). We found
that participants' average fingerprints tended to remain relatively stable on
early lists, and exhibited a ``jump'' to another stable state on later lists.
The sizes of these jumps varied somewhat across conditions (the Euclidean
distances between fingerprints in their original high dimensional spaces are
displayed in the bottom panels). We also averaged the fingerprints across early
and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations on
those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visiual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

(Fig.~\ref{fig:clusteringfig:clustering-scatterplots}E), this trend held

@@

(Fig.~\ref{fig:clusteringfig:clustering-scatterplots}F), the results were more

@@

    \caption{\textbf{Feature clustering carryover effects.} \textbf{A.} Feature
    clustering scores for ordder manipulation (early) versus randomly ordered
    (late) lists. \textbf{B.} Accuracy differences (on early versus late lists)
    versus feature clustering ``carryover'' (defined as the differences between
    the average clustering scores on early and late lists). \textbf{C.}
    Temporal clustering differences (on early versus late lists) versus feature
    clustering carryover. \textbf{All panels.} Each dot in the main
    scatterplots denotes the average scores for one participant. The colored
    regression lines are computed across participants. The inset displays
    condition-averaged results, where each dot reflects a single condition and
    the regression line is computed across experimental conditions. All error
    ribbons denote bootstrap-estimated 95\% confidence intervals.}
    \label{fig:clustering-carryover}

@@

The order manipulations we examined also appeared to induce, in some cases, a
tendancy to ``clump'' similar words within a list. This was most apparent on
semantically ordered lists, where the probability of initiating recall with a
given word seemed to follow groupings defined by feature changepoints.

>>>>

Participants were assigned to experimental conditions based loosely on their
date of participation. (This aspect of our procedure helped us to more easily
synchronize the experiment databases across multiple testing computers.) Of the
490 participants who opted to fill out the demographics survey, reported ages
ranged from 17 to 31 years (mean: 19.1 years; standard deviation: 1.356 years).
A total of 318 participants reported their gender as female, 170 as male, and
two participants declined to report their gender. A total of 442 participants
reported their ethnicity as ``not Hispanic or Latino,'' 39 as ``Hispanic or
Latino,'' and nine declined to report their ethnicity. Participants reported
their races as White (345 participants), Asian (120 participants), Black or
African American (31 participants), American Indian or Alaska Native (11
participants), Native Hawaiian or Other Pacific Islander (four participants),
Mixed race (three participants), Middle Eastern (one participant), and Arab
(one participant). A total of five participants declined to report their race.
We note that several participants reported more than one of racial category.
Participants reported their highest degrees achieved as ``Some college'' (359
participants), ``High school graduate'' (117 participants), ``College
graduate'' (seven participants), ``Some high school'' (five participants),
``Doctorate'' (one participant), and ``Master's degree'' (one participant). A
total of 482 participants reported no reading impairments, and eight reported
having mild reading impairments. A total of 489 participants reported having
normal color vision and one participant reported that they were red-green color
blind. A total of 482 participants reported taking no prescription medications
and having no recent injuries; four participants reported having ADHD, one reported
having dyslexia, one reported having allergies, one reported a recently torn
ACL/MCL, and one reported a concussion from several months prior. The
participants reported consuming 0 -- 3 cups of coffee prior to the testing
session (mean: 0.32 cups; standard deviation: 0.58 cups). Participants reported
their current level of alertness, and we converted their responses to numerical
scores as follows: ``very sluggish'' (-2), ``a little sluggish'' (-1),
``neutral'' (0), ``a little alert'' (1), and ``very alert'' (2). Across all
participants, the full range of alertness levels were reported (range: -2 -- 2;
mean: 0.35; standard deviation: 0.89).

@@

Our experiment is a variant of the classic free recall paradigm that we term
\textit{feature-rich free recall}. In feature-rich free recall, participants
study 16 lists, each comprised of 16 words that vary along a number of stimulus
dimensions (Fig.~\ref{fig:exp}). The stimulus dimensions include two semantic
features related to the \textit{meanings} of the words (semantic category,
referent object size), two lexicographic features related to the
\textit{letters} that make up the words (word length in number of letters,
identity of the word's first letter), and two visual features that are
independent of the words themselves (text color, presentation location). Each
list contains four words from each of four different semantic categories and
two object sizes; all other stimulus features are randomized. After studying
each list, the participant attempts to recall as many words as they can from
that list, in any order they choose. Because each individual word is associated
with several well-defined (and quantifiable) features, and because each list
incorporates a diverse mix of feature values along each dimension, this allows
us to estimate which features participants are considering or leveraging in
organizing their memories.

@@

We assigned the categorized words into a total of 16 lists with several
constraints. First, we required that each list contained words from exactly 4
unique categories, each with exactly 4 exemplars from each category. Second, we
required that (across all words on the list) at least one instance of both
object sizes were represented. On average, each category was represented in
4.27 lists (standard deviation: 1.16 lists). Aside from these two constraints,
we assigned each word to a unique list. After random assignment, each list
contained words with an average of 11.13 unique starting letters (standard
deviation: 1.15 letters) and an average word length of 6.17 letters (standard
deviation: 0.34 letters).

@@

To select a random font color for each word, we drew three integers uniformly
and at random from the interval $\left[0, 255\right]$, corresponding to the red (r),
green (g), and blue (b) color channels for that word. To assign random
presentation locations to each word, we selected two floating point numbers
uniformly at random (one for the word's horizontal $x$ coordinate and the other
for its vertical $y$ coordinate). The bounds of these coordinates were selected
to cover the entire visible area of the display without cutting off any part of
the words. The words were shown on 27~in (diagonal) Retina 5K iMac displays
(resolution: 5120 $\times$ 2880 pixels).

@@

We used four ``control'' conditions to evaluate and explore participants'
baseline behaviors. We also used performance on these control conditions to
help interpret performance in other ``manipulation'' conditions. Two control
conditions served as ``anchor points.'' In the first anchor point condition,
which we call the \textit{feature rich} condition, we randomly shuffled the
presentation order (independently for each participant) of the words on each
list. In the second anchor point condition, which we call the \textit{reduced}
condition, we randomized word presentations as in the feature rich condition.
However, rather than assigning each word a random color and location, we
instead displayed all of the words in black and at the center of the screen.

@@

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique lists for each participant.
First, we choose a word uniformly at random from the set of candidates. Next,
we compute the distances between the chosen word's feature(s) and the
corresponding feature(s) of all yet-to-be-presented words. Third, we convert
these distances (between the previously presented word's feature values, $a$,
and the candidate word's feature values, $b$) to similarity scores:
\begin{equation} \mathrm{similarity}(a, b) = \exp\{-\tau \cdot
\mathrm{distance}(a, b)\}, \label{eqn:similarity} \end{equation} where $\tau =
1$ in our implementation. We note that increasing the value of $\tau$ would
amplify the influence of similarity on order, and decreasing the value of
$\tau$ would diminish the influence of similarity on order. Also note that this
approach requires $\tau > 0$. Finally, we computed a set of normalized
similarity values by dividing the similarities by their sum: \begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, \end{equation} where in the
denominator, $i$ takes on each of the $n$ feature values of the to-be-presented
words. The resulting set of normalized similarity scores sums to one.

@@

\caption{\textbf{Generating stochastic feature-sorted lists.} For a given
feature dimension (e.g., color), we compute the similarity
(Eqn.~\ref{eqn:similarity}) between the feature value(s) of the previous item,
$x$, and all yet-to-be-presented items ($a$ -- $g$). Next, we normalize these
similarity scores so that they sum to one. We lay, in sequence, a set of
``sticks,'' one for each candidate item, whose lengths are equal to these
normalized similarity scores. Note that the combined lengths of these sticks is
one. To select the next to-be-presented item, we draw a random number, $r$,
from the uniform distribution bounded between 0 and 1 (inclusive). The identity
of the next item is given by the stick adjacent to an indicator that moves
distance $r$ (starting from 0) along the sequence of sticks. In this case, the
next to-be-presented item is $e$. Note that each item's chances of selection is
proportional to its similarity to the previous item, along the given feature
dimension.} \label{fig:stick}

@@

Suppose that two lists contain unequal numbers of items of each size. For
example, suppose that list $A$ contains all ``large'' items, whereas list $B$
contains an equal mix of ``large'' and ``small'' items. For a participant
recalling list $A$, any correctly recalled item will necessarily match the size
of the previous correctly recalled item. In other words, successively recalling
several list $A$ items of the same size is essentially meaningless, since
\textit{any} correctly recalled list $A$ word will be large. In contrast,
successively recalling several list $B$ items \textit{could} be meaningful,
since (early in the recall sequence) the yet-to-be-recalled items come from a
mix of sizes. However, once all of the small items on list $B$ have been
recalled, the best possible next matching recall will be a large item. And all
subsequent correct recalls must also be large items-- so for those later
recalls it becomes difficult to determine whether the participant is
successively recalling large items because they are organizing their
memories according to size, or (alternatively), whether they are simply
recalling the yet-to-be-recalled items in a random order. In general, the
precise order and blend of feature values expressed in a given list, the orders
and numbers of correct recalls a participant makes, the number of intervening
presentation positions between successive recalls, and so on, can all affect
the range of clustering scores that are possible to observe for a given list.
An uncorrected clustering score therefore conflates participants' actual
memory organization with other ``nuisance'' factors.

@@

We define each participant's \textit{memory fingerprint} as the set of their
permutation-corrected clustering scores across all dimensions we tracked in our
study, including their six feature-based clustering scores (category, size,
length, first letter, color, and location) and their temporal clustering score.
Conceptually, a participant's memory fingerprint describes their tendency to
order in their recall sequences (and, presumably, organize in memory) the
studied words along each dimension. To obtain stable estimates of these
fingerprints for each participant, we averaged clustering scores across lists.
We also tracked and characterized how participants' fingerprints changed across
lists (e.g.,
Figs.~\ref{fig:fingerprint-trajectories},~\fingerprintTrajectoryRandom).

@@

The presentation orders of some lists in the adaptive condition of our
experiment (see \textit{Adaptive condition}) were sorted according to
participants' \textit{current} memory fingerprint, estimated using all of the
lists they had studied up to that point in the experiment. Because our
experiment incorporated a speech-to-text component, all of the behavioral data
for each participant could be analyzed just a few seconds after the conclusion
of the recall intervals for each list. We used the \texttt{Quail} Python
package~\citep{HeusEtal17} to apply speech-to-text algorithms to the
just-collected data, aggregate the data for the given participant, and estimate
the participant's memory fingerprint using all of their available data up to
that point in the experiment. Two aspects of our implementation are worth
noting. First, because memory fingerprints are computed independently for each
list and then averaged across lists, the already-computed memory fingerprints
for earlier lists could be cached and loaded as needed in future computations.
This meant that our computations pertaining to updating our estimate of a
participant's memory fingerprint only needed to consider data from the most
recent list. Second, each element of the null distributions of uncorrected
fingerprint scores (see \textit{Permutation-corrected feature clustering
scores}) could be estimated independently from the others. This enabled us to
make use of the testing computers' multi-core CPU architectures by elements of
the null distributions in batches of eight (i.e., the number of CPU cores on
each testing computer). Taken together, we were able to compress the relevant
computations into just a few seconds of computing time. The combined processing
time for the speech-to-text algorithm, fingerprint computations, and
permutation-based ordering procedure (described next) easily fit within the
inter-list intervals, where participants paused for a self-paced break before
moving on to study and recall the next list.

@@

Probability of first recall curves~\citep{AtkiShif68, PostPhil65, WelcBurn24}
reflect the probability that an item will be recalled first, as a function of
its serial position during encoding. To carry out this analysis, we initialized
(for each participant) a number-of-lists (16) by number-of-words-per-list (16)
matrix of zeros. Then, for each list, we found the index of the word that was
recalled first, and we filled in that position in the matrix with a 1. Finally,
we averaged over the rows of the matrix to obtain a 1 by 16 array of
probabilities, for each participant. We used an analogous procedure to compute
probability of $n$\textsuperscript{th} recall curves for each participant.
Specifically, we filled in the corresponding matrices according to the
$n$\textsuperscript{th} recall on each list that each participant made. When a
given participant had made fewer than $n$ recalls for a given list, we simply
excluded that list from our analysis when computing that participant's
curve(s).

@@

We used the distances between feature values for successively presented words
(see \textit{Defining feature-based distances}) to estimate ``event
boundaries'' where the feature values changed more than
usual~\citep{EzzyDava11, MannEtal16,RadvCope06, SwalEtal09, SwalEtal11,
DuBrDava16}. For each list, for each feature dimension, we computed the
distribution of distances between the feature values for successively presented
words. We defined event boundaries (e.g., Fig.~\ref{fig:recall-dynamics}B) as
occurring between any successive pair of words whose distances along the given
feature dimension were greater than one standard deviation above the mean for
that list. Note that, because event boundaries are defined for each feature
dimension, each individual list may contain several sets of event boundaries,
each at different moments in the presentation sequence (depending on the
feature dimension of interest).

@@

While holding the set of words (and the assignments of words to lists)
constant, we manipulated two aspects of participants' experiences of studying
each list. We sought to understand the effects of these manipulations on
participants' memories for the studied words. First, we added two additional
sources of visual variation to the individual word presentations: font color
and onscreen location. Importantly, these visual features were independent of
the meaning or semantic content of the words (e.g., word category, size of the
referent, etc.) and of the lexicographic properties of the words (e.g., word length,
first letter, etc.). We wondered whether this additional word-independent information
might facilitate recall (e.g., by providing new potential ways of organizing or
retrieving memories of the studied words) or impair recall (e.g., by
distracting participants with irrelevant information). Second, we manipulated
the orders in which words were studied (and how those orderings changed over
time). We wondered whether presenting the same list of words with different
appearances (e.g., by manipulating font size and onscreen location) or in
different orders (e.g., sorted along one feature dimension versus another)
might serve to influence how participants organized their memories of the
words. We also wondered whether some order manipulations might be temporally
``sticky'' by influencing how \textit{future} lists were remembered.

@@

manipulation conditions was indistinguishable from performance on the early

@@

the semantically ordered conditions exhibited stronger temporal clustering on

@@

conditions. Since results were similar across both semantic conditions

@@

When we closely examined the sequences of words participants recalled in early
order manipulated lists (Fig.~\ref{fig:recall-dynamics}A, top panel), we
noticed several differences from the dynamics of participants' recalls of
randomly ordered lists (Figs.~\dynamicsRandom,~\recallInit). One striking difference is that
participants in the category condition (dark purple curves,
Fig.~\ref{fig:recall-dynamics}) most often initiated recall with the
fourth-from-last item (\textit{Recall initiation}, top left panel), whereas
participants who recalled randomly ordered lists tended to initiate recall with
either the first or last list items (Fig.~\dynamicsRandom, top left panel). We
hypothesized that the participants might be ``clumping'' their recalls into
groups of items that shared category labels. Indeed, when we compared the
positions of feature changes in the study sequence
(Fig.~\ref{fig:recall-dynamics}B; see \textit{Identifying event boundaries})
with the positions of items participants recalled first, we noticed a striking
correspondence in both semantic conditions. Specifically, on category-ordered
lists, the category labels changed every four items on average (dark purple
peaks in Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to
display an increased tendency (relative to other order manipulation and random
conditions) to initiate recall of category-ordered lists with items whose study
positions were integer multiples of four. Similarly, for size-ordered lists,
the size labels changed every eight items on average (light purple peaks in
Fig.~\ref{fig:recall-dynamics}B), and participants also seemed to display an
increased tendency to initiate recall of size-ordered lists with items whose
study positions were integer multiples of eight. A second striking difference
is that participants in the category condition exhibited a much steeper lag-CRP
(Fig.~\ref{fig:recall-dynamics}A, top middle panel) than participants in other
conditions. (This is another expression of participants' increased tendencies
to temporally cluster their recalls on category-ordered lists, as we reported
above.) Taken together, these order-specific idiosyncrasies suggest a
hierarchical set of influences on participants' memories. At longer timescales,
``event boundaries'' (to use the term loosely) can be induced across lists by
adding or removing irrelevant visual features. At shorter timescales, ``event
boundaries'' can be induced across items (within a single list) by adjusting
how item features change throughout the list.

@@

We also looked for more subtle group-level patterns. For example, perhaps
sorting early lists by one feature dimension could affect how participants
cluster \textit{other} features (on early and/or late lists) as well. We
defined participants' \textit{memory fingerprints} as the set of temporal and
feature clustering scores. A participant's memory fingerprint describes how
they tend to retrieve memories of the studied items, perhaps searching through
several feature spaces (or along several representational dimensions). To gain
insights into the dynamics of how participants' clustering scores tended to
change over time, we computed the average (across participants) fingerprint
from each list, from each order manipulation condition
(Fig.~\ref{fig:fingerprint-trajectories}). We projected these fingerprints into
a two-dimensional space to help visualize the dynamics (top panels; see
\textit{Computing low-dimensional embeddings of memory fingerprints}). We found
that participants' average fingerprints tended to remain relatively stable on
early lists, and exhibited a ``jump'' to another stable state on later lists.
The sizes of these jumps varied somewhat across conditions (the Euclidean
distances between fingerprints in their original high dimensional spaces are
displayed in the bottom panels). We also averaged the fingerprints across early
and late lists, respectively, for each condition
(Fig.~\ref{fig:fingerprint-trajectories}B). We found that participants'
fingerprints on early lists seem to be influenced by the order manipulations on
those lists (see the locations of the circles in
Fig.~\ref{fig:fingerprint-trajectories}B). There also seemed to be some
consistency across different features within a broader type. For example, both
semantic feature conditions (category and size; purple markers) diverge in a
similar direction from the group; both lexicographic feature conditions (length
and first letter; yellow markers) diverge in a similar direction; and both
visual conditions (color and location; green) also diverge in a similar
direction. But on late lists, participants' fingerprints seem to return to a
common state that is roughly shared across conditions (i.e., the stars in that
panel are clumped together).

@@

(Fig.~\ref{fig:clustering-scatterplots}E), this trend held

@@

(Fig.~\ref{fig:clustering-scatterplots}F), the results were more

@@

    \caption{\textbf{Feature clustering carryover effects.} \textbf{A.} Feature
    clustering scores for order manipulation (early) versus randomly ordered
    (late) lists. \textbf{B.} Accuracy differences (on early versus late lists)
    versus feature clustering ``carryover'' (defined as the differences between
    the average clustering scores on early and late lists). \textbf{C.}
    Temporal clustering differences (on early versus late lists) versus feature
    clustering carryover. \textbf{All panels.} Each dot in the main
    scatterplots denotes the average scores for one participant. The colored
    regression lines are computed across participants. The inset displays
    condition-averaged results, where each dot reflects a single condition and
    the regression line is computed across experimental conditions. All error
    ribbons denote bootstrap-estimated 95\% confidence intervals.}
    \label{fig:clustering-carryover}

@@

The order manipulations we examined also appeared to induce, in some cases, a
tendency to ``clump'' similar words within a list. This was most apparent on
semantically ordered lists, where the probability of initiating recall with a
given word seemed to follow groupings defined by feature change points.
